Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-14 analysis/paper/paper.qmd
updates of data and text

library(data.table)
library(NISTunits)
library(circular)
library(ggpubr)
library(ggalt)
library(ggplot2)
library(stringr)
library(ggtern)
library(dplyr)
library(here)
library(tidyverse)
library(tidyr)
library(ggrepel)
library(here)

@@

df <- tibble(
  material = c("Lava", "Dolomite", "Chert", "Other"),
  spheroids = c(10, 2, 0, 0),
  cores = c(61, 21, 4, 8),
  Polys_and_Multi_cores = c(20, 5, 3, 2),
  total = c(637, 209, 53, 56),
)

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). Among these, 6 spheroids and 6 subspheroids were identified; together forming a spheroid group of 12 pieces, accounting for 1.3% of the total lithic assemblage. They are all faceted spheroids with intensive flake scars on the surface. Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.

@@

As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020]. Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020]. The resulting coefficients were normalized for scale-invariant comparison across models.

@@

library(readxl)
library(Hmisc)       
library(ggcorrplot)  
library(reshape2)

@@

selected_names <- c("Scar.number", "Removal.ratio", "SDI",
                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")
selected_cols <- data[, selected_names]

@@

# edge angle ~ sphericity
val4 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Sphericity", ]
cor_angle_sph <- sprintf("%.2f", val4$cor)
p_angle_sph <- pretty_print_sci(val4$p)

# edge angle ~ SHE
val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SHE", ]
cor_angle_she <- sprintf("%.2f", val5$cor)
p_angle_she <- pretty_print_sci(val5$p)

@@

# Edge angle ~ Scar number
val_edge_scar <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Scar.number", ]
cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
p_edge_scar <- pretty_print_sci(val_edge_scar$p)

# Edge angle ~ Removal ratio
val_edge_removal <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Removal.ratio", ]
cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
p_edge_removal <- pretty_print_sci(val_edge_removal$p)

@@

ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
  geom_tile(color = "gray") +
  scale_fill_gradient2(low = "#5e3c99", mid = "white", high = "#e66101", midpoint = 0) +
  geom_text(aes(label = cor_label), size = 3) +
  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "Correlation", x = "", y = "")
```

>>>>

library(data.table)
library(NISTunits)
library(circular)
library(ggpubr)
library(ggalt)
library(ggplot2)
library(stringr)
library(ggtern)
library(dplyr)
library(here)
library(tidyverse)
library(tidyr)
library(ggrepel)
library(here)
library(readxl)
library(Hmisc)       
library(ggcorrplot)  
library(reshape2)

@@

df <- tibble(
  material = c("Lava", "Dolomite", "Chert", "Other"),
  spheroids = c(11, 2, 0, 0),
  cores = c(60, 21, 4, 8),
  Polys_and_Multi_cores = c(19, 5, 3, 2),
  total = c(637, 209, 53, 56),
)

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). Among these, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 pieces, accounting for 1.3% of the total lithic assemblage. They are all faceted spheroids with intensive flake scars on the surface. Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.

@@

As shown in the @fig-SPHARM-steps, following the approach by @Medyukhina2020 and @Link2024, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ). Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@wieczorek2018shtools]. The resulting coefficients were normalized for scale-invariant comparison across models.

@@

@@

selected_names <- c("Scar.number", "Removal.ratio", "SDI",
                    "Mean.edge.angle", "Curvature", "Sphericity", "SHE")
selected_cols <- data[, selected_names]

@@

# edge angle ~ sphericity
val4 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Sphericity", ]
cor_angle_sph <- sprintf("%.2f", val4$cor)
p_angle_sph <- pretty_print_sci(val4$p)

# edge angle ~ SHE
val5 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "SHE", ]
cor_angle_she <- sprintf("%.2f", val5$cor)
p_angle_she <- pretty_print_sci(val5$p)

@@

# Edge angle ~ Scar number
val_edge_scar <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Scar.number", ]
cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
p_edge_scar <- pretty_print_sci(val_edge_scar$p)

# Edge angle ~ Removal ratio
val_edge_removal <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Removal.ratio", ]
cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
p_edge_removal <- pretty_print_sci(val_edge_removal$p)

@@

p <- 
  ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
    geom_tile(color = "gray") +
    scale_fill_gradient2(low = "#5e3c99", mid = "white", high = "#e66101", midpoint = 0) +
    geom_text(aes(label = cor_label), size = 3) +
    geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(fill = "Correlation", x = "", y = "")

ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
       plot = p, width = 8, height = 6, dpi = 300)
```
